---
title: "Chikungunya Host and Vector SDM Bundle Summary"
output:
  html_document:
    toc: true
    toc_float: true
    toc_depth: 2
    theme: flatly
    highlight: tango
    fig_width: 10
    fig_height: 6
editor_options:
  chunk_output_type: inline
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(
  echo = FALSE,
  message = FALSE,
  warning = FALSE,
  fig.width = 10,
  fig.height = 6
)

has_path <- function(x) {
  is.character(x) && length(x) > 0 && !is.na(x[[1]]) && nzchar(x[[1]]) &&
    !identical(toupper(trimws(x[[1]])), "NA")
}

parent_dirs <- function(path) {
  if (!has_path(path)) return(character(0))
  path <- normalizePath(path[[1]], mustWork = FALSE)
  out <- character(0)
  repeat {
    out <- c(out, path)
    parent <- dirname(path)
    if (identical(parent, path)) break
    path <- parent
  }
  unique(out)
}

find_bundle_root <- function(...) {
  candidates <- unique(unlist(lapply(list(...), parent_dirs), use.names = FALSE))
  default_delivery_root <- "/Volumes/LaCie/new_global_maxent/sdms/delivery/chikungunya_vector_sdm_delivery_20260609"
  if (dir.exists(default_delivery_root)) candidates <- c(candidates, default_delivery_root)

  for (candidate in unique(candidates)) {
    if (file.exists(file.path(candidate, "manifest.csv"))) return(candidate)
  }

  stop("Could not find manifest.csv. Open this Rmd from the delivery bundle folder, or set `bundle_root` manually.", call. = FALSE)
}

notebook_path <- tryCatch({
  if (requireNamespace("knitr", quietly = TRUE)) knitr::current_input(dir = TRUE) else ""
}, error = function(e) "")

if (!has_path(notebook_path) && requireNamespace("rstudioapi", quietly = TRUE)) {
  notebook_path <- tryCatch(rstudioapi::getActiveDocumentContext()$path, error = function(e) "")
}

bundle_root <- find_bundle_root(notebook_path, getwd())
source(file.path(bundle_root, "explore_chikungunya_vector_sdm_bundle.R"))

render_table <- function(dat, digits = 3) {
  dat <- as.data.frame(dat)
  numeric_cols <- vapply(dat, is.numeric, logical(1))
  dat[numeric_cols] <- lapply(dat[numeric_cols], round, digits = digits)

  knitr::kable(
    dat,
    format = "html",
    escape = TRUE,
    table.attr = "class='table table-striped table-hover table-condensed'"
  )
}

drop_geometry_if_needed <- function(dat) {
  if (inherits(dat, "sf") && requireNamespace("sf", quietly = TRUE)) {
    return(sf::st_drop_geometry(dat))
  }
  dat
}

include_saved_map <- function(species, with_occurrences = TRUE) {
  path <- map_path(species, with_occurrences = with_occurrences)
  if (!file.exists(path) && with_occurrences) path <- map_path(species, with_occurrences = FALSE)
  if (!file.exists(path)) return(invisible(NULL))

  knitr::include_graphics(path)
}
```

This static report summarizes the delivered Chikungunya host and vector SDM bundle. It is intended for quick review and sharing; use `explore_chikungunya_vector_sdm_bundle.Rmd` for interactive filtering, downloads, generated code, and dynamic map selection.

## Disease Summary

```{r disease-summary}
disease_info <- disease_overview()
country_metrics <- pilot_country_metrics()

disease_summary_table <- tibble::tibble(
  field = c(
    "Disease",
    "Analysis unit",
    "Pathogen",
    "Pathogen taxid",
    "Family",
    "Transmission guild",
    "Range-limiting layer",
    "Host SDMs available",
    "Vector SDMs available",
    "Countries/territories with evidence",
    "GenBank-supported countries/territories",
    "WHO DON-supported countries/territories",
    "Countries/territories supported by both"
  ),
  value = c(
    disease_info$readiness_disease_name,
    disease_info$analysis_unit_label,
    disease_info$pathogen_species_name,
    disease_info$pathogen_taxid,
    disease_info$family,
    disease_info$guild,
    disease_info$range_limiting_layer,
    disease_info$host_sdm_species_available,
    disease_info$vector_sdm_species_available,
    country_metrics$countries_or_territories,
    country_metrics$genbank_supported,
    country_metrics$who_don_supported,
    country_metrics$supported_by_both
  )
)

render_table(disease_summary_table)
```

## Country Evidence

```{r country-evidence}
country_map_packages_missing <- missing_country_map_packages()
country_summary_table <- if (length(country_map_packages_missing) == 0) {
  pilot_country_summary(map_ready = TRUE) |>
    drop_geometry_if_needed()
} else {
  pilot_country_summary(map_ready = FALSE)
}

country_source_counts <- country_summary_table |>
  dplyr::count(source_group, name = "countries_or_territories") |>
  dplyr::arrange(dplyr::desc(countries_or_territories))

render_table(country_source_counts)
```

```{r country-map, fig.height=6.5}
if (length(country_map_packages_missing) == 0) {
  print(pilot_country_map_plot())
} else {
  cat("Install required map packages to draw the country map: ", paste(country_map_packages_missing, collapse = ", "), "\n", sep = "")
}
```

## Bundle Summary

```{r bundle-summary}
sdm_status <- sdm_summary()
sdm_index <- sdms(available_only = FALSE)

render_table(sdm_status)
```

```{r availability-counts}
availability_counts <- tibble::tibble(
  category = c("Delivered host SDMs", "Delivered vector SDMs", "Diagnostic vector SDMs"),
  n = c(
    nrow(hosts(sdm_available = TRUE)),
    nrow(vectors(sdm_available = TRUE)),
    nrow(diagnostic_sdms())
  )
)

render_table(availability_counts)
```

## Delivered Host SDMs

```{r host-sdms}
host_table <- filter_hosts_for_bundle(sdm_availability = "available") |>
  dplyr::select(-local_model_path, -local_prediction_tif) |>
  dplyr::arrange(species_name)

render_table(host_table)
```

## Delivered Vector SDMs

```{r vector-sdms}
vector_table <- filter_vectors_for_bundle(sdm_availability = "available") |>
  dplyr::select(-local_model_path, -local_prediction_tif) |>
  dplyr::arrange(delivery_group, species_name)

render_table(vector_table)
```

## Diagnostic Vectors

These three vector SDMs are included for inspection but are not recommended as clean production models without manual review.

```{r diagnostic-vectors}
diagnostic_table <- diagnostic_sdms() |>
  dplyr::select(
    species_name,
    delivery_group,
    model_quality,
    retained_models,
    min_boyce,
    max_boyce,
    max_test_auc,
    max_max_tss
  ) |>
  dplyr::arrange(species_name)

render_table(diagnostic_table)
```

## Example Prediction Maps

These are saved maps from the delivered bundle. The full set of `.rds` model paths and `.tif` raster paths is available through the interactive notebook downloads or the helper functions.

### Aedes aegypti

```{r map-aedes-aegypti, fig.height=6.5}
include_saved_map("Aedes aegypti", with_occurrences = TRUE)
```

### Aedes albopictus

```{r map-aedes-albopictus, fig.height=6.5}
include_saved_map("Aedes albopictus", with_occurrences = TRUE)
```

### Opifex fuscus

```{r map-opifex-fuscus, fig.height=6.5}
include_saved_map("Opifex fuscus", with_occurrences = TRUE)
```

### Aedes africanus

```{r map-aedes-africanus, fig.height=6.5}
include_saved_map("Aedes africanus", with_occurrences = TRUE)
```
